---
title: "Malaria Dataset Analysis"
output: html_document
---

```{r setup, include=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This document contains my malaria dataset analysis. 

Dataset Used: <https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-11-13>

## Import relevant libraries

```{r}
#import libraries
#if you havent install the relevant package, please install them.
#install.packages('dplyr')
library(dplyr)
#install.packages('ggplot2')
library(ggplot2)
```
-----------------------------------------------------------------------------------

## Malaria_inc dataset analysis
Answer I would like to find out from this dataset analysis:

1. Which countries have the highest malaria cases? (top 10)

2. From the top 10 countries from (1), are there any improvement over the years?

```{r reading malaria_incidence dataset}
#reading malaria_incidence dataset
malaria_inc <-read.csv("malaria_inc.csv")
names(malaria_inc)
str(malaria_inc)
```
Observation: 

1) The dataset only contains 4 years - 2000,2005,2010,2015. To keep the analysis consistent, I will use only these 4 values for the rest of my analysis.

2) The first column name is called "Entity" which is basically the countries. The last column name is "Incidence.of.malaria..per.1.000.population.at.risk...per.1.000.population.at.risk." which is the number of cases per 1000 population. (For better readibility, I will rename the two columns.)

```{r rename malaria_inc dataset}
#rename first column to country and last column to cases for better readibility
malaria_inc<-malaria_inc %>% setNames(c("country", "code", "year", "cases"))
```

```{r finding top 10 highest cases}
#find out the top 10 countries that have the highest cases for the span of 15 years
worst_inc <- malaria_inc %>%
  group_by(country) %>%
  summarise(cases = sum(cases)) %>%
  arrange(desc(cases))%>%
  head(10)
```
Observation: 

The top 10 countries with the highest cases for the span of 15 years are Burkina Faso,Turkey, Cote d'Ivoire, Togo, Mali, Nigeria, Niger, Sierra Leone, Democratic Republic of Congo and Guinea.

```{r filtered and plot for top10 countries with highest cases}
#filtered dataset based on the the top 10 countries with the highest cases
worst_inc_year<-malaria_inc %>%
  group_by(year,country) %>%
  summarise(cases = sum(cases)) %>%
  arrange(country,year) %>%
  filter(country %in% worst_inc$country)

#plot to check whether there are any improvements over the years-decrease in cases
ggplot(data=worst_inc_year,aes(x=year,y=cases,color = country))+geom_line()
```

Observation: 

For most of the countries, there were not much significant improvements in the reduction of cases. The only country that have significant improvement is Turkey.

-----------------------------------------------------------------------------------

## Malaria_death dataset analysis

Answer I would like to find out from this dataset analysis:

1. Which countries have the highest malaria deaths? (top 10) Are the results the same as those countries with the highest cases?

2. From the top 10 countries from (1), are there any improvement over the years?

```{r reading malaria_deaths dataset}
#reading malaria_deaths dataset
malaria_deaths <-read.csv("malaria_deaths.csv")
names(malaria_deaths)
str(malaria_deaths)
```

Observations:

1) The first column name is called "Entity" which is basically the countries. The last column name is "Deaths...Malaria...Sex..Both...Age..Age.standardized..Rate...per.100.000.people." which is the number of deaths per 100,000 people.(For better readibility, I will rename the two columns.)

```{r rename malaria_death dataset}
#rename first column to country and last column to deaths for better readibility
malaria_deaths<-malaria_deaths %>% setNames(c("country", "code", "year", "deaths"))
```

```{r filtered dataset only containing the selective year}
#filter dataset to only contains Years = 2000,2005,2010,2015 to ensure the analysis is consistent for comparison 
malaria_deaths_filtered_year <- malaria_deaths %>%
  filter(year==2000|year == 2005|year==2010|year==2015)
```

```{r worst_death rate countries}
#find out which countries have the worst death rates
worst_countries <- malaria_deaths_filtered_year %>%
  group_by(country) %>%
  summarise(deaths = sum(deaths)) %>%
  arrange(desc(deaths))%>%
  head(10)
```

Observations: 

Comparing the top 10 countries with the highest death rates and highest case rates, 8 out of 10 are the same.This proves that with higher cases rate, there will be higher deaths. 

The top 10 countries with the highest death rates are Sierra Leone, Burkina Faso,Equatorial Guinea, Cote d'Ivoire, Niger, Nigeria, Uganda, Democratic Republic of Congo, Mali, Cameroon

```{r filtered based on the top 10 countries with highest death rate}
#filtered dataset based on the the top 10 countries with the highest deaths rate
worst_countries_year<-malaria_deaths_filtered_year %>%
  group_by(year,country) %>%
  summarise(deaths = sum(deaths)) %>%
  arrange(country,year) %>%
  filter(country %in% worst_countries$country)

#plot to check whether there are decrement in death rates over the years
ggplot(data=worst_countries_year,aes(x=year,y=deaths,color = country))+geom_line() +facet_wrap(~country)
```

Observation: 

Most of the countries have a downward trends meaning over the span of 15 years, the death rates have been decreasing. There is only 1 country that have insignificant improvement - Mali.

-----------------------------------------------------------------------------------
## Malaria_death_age dataset analysis
Answer I would like to find out from this dataset analysis:

1. For the top 10 countries with the highest death rates, which age group have a higher percentage of deaths?


```{r reading malaria_deaths_age dataset}
#reading malaria_deaths_age dataset
malaria_deaths_age <-read.csv("malaria_deaths_age.csv")
names(malaria_deaths_age)
str(malaria_deaths_age)
```

Observations:

1) The first column name is called "x" which is the running index. The second column name is called "Entity" which is basically the countries.(For better readibility, I will rename the two columns.)

```{r rename malaria_deaths_age cols}
#rename the columns
malaria_deaths_age<-malaria_deaths_age %>% setNames(c("index","country", "code", "year","age_group", "deaths"))
```

```{r filtered malaria_deaths_age dataset to only contains selective years}
#filter dataset to only contains Years = 2000,2005,2010,2015 to ensure the analysis is consistent for comparison 
malaria_deaths_age_filtered_year<-malaria_deaths_age %>%
  filter(year==2000|year == 2005|year==2010|year==2015)
```

```{r filtered based on top 10 highest death rate}
#filter based on the top 10 countries with the highest death rate.As this dataset is related to death rates, I used the top 10 countries that has the highest death rates(as produced by the second analysis). 
worst_countries_year_age<-malaria_deaths_age_filtered_year %>%
  filter(country %in% worst_countries$country)

```

```{r}
#plot to check the age group that are more vulnerable
ggplot(worst_countries_year_age, aes(fill=age_group, y=deaths, x=year)) + 
    geom_bar(position="fill", stat="identity")+ facet_wrap(~country)
```

Observations:

For all the 10 countries with the highest death rates, the highest death rates age group is "Under 5". Therefore, it is important to focus any malaria treatment to children to reduce the death rates as children immunity is lower.


Summary

1) Poorly developed countries like Africa have the highest chance to suffer from Malaria. 

2) The most vulnerable age group is children under 5.

3) Despite many efforts to help reduce the spread of Malaria, poorly developed countries are still suffering from Malaria although there is an significant improvement to the death rates.
